---
title: "hypothesis 4"
output:
  pdf_document: default
  html_document: default
date: "2023-06-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = "https://cloud.r-project.org")

library(car)
library(lme4)
library(multcomp)
library(ggplot2)
install.packages("knitr")
install.packages("kableExtra")
library(MASS)
library(emmeans)


library(knitr)
library(kableExtra)

```


```{r, echo=FALSE}
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

```

Hypothesis 4: What is the relationship between the abundance of H.perlevis and biomass corrected habitat (2022 study)**</div></font>
  
```{r, echo=FALSE}
library(ggplot2)
data2022 <- read.csv('Dataa/data2022.csv')
data2022$row_id <- factor(1:nrow(data2022))
data2022 <- data2022[data2022$Abundance.sponges > 0, ]
data2022$Sponge.biomass <- as.numeric(gsub("[^0-9.]", "", data2022$Sponge.biomass))
summarydata2022 <- summary(data2022)
kable(summarydata2022, caption="summary data 2022")

# cleaning up
data2022$ln.biovolume <- log(data2022$Sponge.biomass/data2022$Abundance.sponges)
data2022$log.n <- log(data2022$Abundance.sponges)
data2022$row_id <- factor(1:nrow(data2022))


#mean volume and sd per habitat
sd_biomass <- aggregate(Sponge.biomass ~ Habitat, data2022, FUN = sd)
mean_biomass <- aggregate(Sponge.biomass ~ Habitat, data2022, FUN = mean)
biomass_mean_sd <- merge(mean_biomass,sd_biomass, by = "Habitat")
colnames(biomass_mean_sd) <- c("Habitat", "Mean", "SD")
kable(biomass_mean_sd,caption= "biomass sd and mean")


#mean abundance and SD per habitat
mean_abundance <- aggregate(Abundance.sponges~ Habitat, data2022, FUN = mean)
sd_abundance <- aggregate(Abundance.sponges~ Habitat, data2022, FUN = sd)
abundance_mean_sd <- merge(mean_abundance,sd_abundance, by = "Habitat")
colnames(abundance_mean_sd) <- c("Habitat", "Mean", "SD")
kable(abundance_mean_sd,caption= "Abundance sd and mean")

# highest and lowest volumes and abundance

#Min values and habitat they belong to 

lowest_volume1 <- min(data2022$Sponge.biomass)
lowest_volume_rows1 <- data2022[data2022$Sponge.biomass == lowest_volume1, ]
kable(lowest_volume_rows1, caption= "lowest volume")

highest_volume1 <- max(data2022$Sponge.biomass)
highest_volume_rows1 <- data2022[data2022$Sponge.biomass == highest_volume1, ]
kable(highest_volume_rows1,caption= "highest volume")

lowest_abundance1<- min(data2022$Abundance.sponges)
lowest_abundance_rows1 <- data2022[data2022$Abundance.sponges == lowest_abundance1, ]
kable(lowest_abundance_rows1, caption= "lowest abundance")


highest_abundance1<- max(data2022$Abundance.sponges)
highest_abudance_rows1 <- data2022[data2022$Abundance.sponges == highest_abundance1, ]
kable(highest_abudance_rows1,caption= "highest abundance")


# model (GLM) 
m10 <- glm(Abundance.sponges ~ Habitat*ln.biovolume, family=poisson,data = data2022)
ahh <- overdisp_fun(m10) #overdispersed
kable(ahh,caption= "m10 overdispersion model habitat and log biovolume with interaction")


m10a <- glm.nb(Abundance.sponges ~ Habitat*ln.biovolume,data = data2022) # negative binomial distribution to account for overdispersion
chisq <- drop1(m10a,test='Chisq') #homogeneity of slopes, no need for interaction term
kable(chisq,caption="chisq test on habitat and volume interaction term on abundance")


m10b <- glm.nb(Abundance.sponges ~ Habitat + ln.biovolume,data = data2022)
ohhh<-overdisp_fun(m10b) # not overdispered
kable(ohhh, caption="overdispersion habitat and volume affect on abundance, no interaction term ")
ahhhhhh <- anova(m10b)
kable(ahhhhhh,caption= "anova on m10b, volume and habitat effect on abundance of sponges")


posthoc <- emmeans(m10b, "Habitat")
pairwise <- pairs(posthoc)
kable(pairwise,caption= "tukey posthoc test on m10b using emmeans package ")
ci <- confint(pairwise)
kable(ci,caption= "CI for tukey posthoc")


#plotting
pred <- predict(m10b)
ggplot(data2022,aes(x=ln.biovolume,y=log(Abundance.sponges),color=Habitat)) + geom_point() +
  geom_line(aes(y=pred)) +
  xlab("Log(Biomass)") +
  ylab("Log(Abundance)")

plot(fitted.values(m10b),residuals(m10b,type='deviance'))
lines(lowess(fitted.values(m10b),residuals(m10b,type='deviance')))

#assumptions
qqnorm(residuals(m10b,type='deviance'))
qqline(residuals(m10b,type='deviance'))




