---
title: "Hypothesis 3"
output:
  pdf_document: default
  html_document: default
date: "2023-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = "https://cloud.r-project.org")
library(car)
library(lme4)
library(multcomp)
library(ggplot2)
install.packages("knitr")
install.packages("kableExtra")


library(knitr)
library(kableExtra)
```

```{r echo=FALSE}
prey_ab_vol <- read.csv('Dataa/prey_ab_vol.csv')
prey_ab_vol$n <- prey_ab_vol$N.H..perlevis
prey_ab_vol$row_id <- factor(1:nrow(prey_ab_vol))
prey_ab_vol$Date <- factor(prey_ab_vol$Date)
prey_ab_vol <- prey_ab_vol[, colSums(is.na(prey_ab_vol)) == 0]

#cleaning up names 
# Rename specific columns by name
old_names <- c("N.quadrat", "N.H..perlevis","Volumen.Hym..mL."," Date","row_id")  # Example: column names to be renamed
new_names <- c("Quadrant number ", "Abundance (n)","Volume (mL)","Date","row_id")  # Example: new names for the columns

for (i in 1:length(old_names)) {
  colnames(prey_ab_vol)[colnames(prey_ab_vol) == old_names[i]] <- new_names[i]
}
prey_ab_vol$log.v <-log(prey_ab_vol$`Volume (mL)`)
```
Hypothesis 3: Is there a relationship between average volume of H.perlevis per month and pH, salinity and temperature averages for the months**</div></font>
  
```{r, echo=FALSE}
TpHSal <- read.csv('Dataa/TempsalinitypH.csv')
empty_columns <- which(colSums(is.na(TpHSal)) == nrow(TpHSal))
TpHSal <- subset(TpHSal, select = -empty_columns)


# summary statistics 
summaryTpHSal <- summary(TpHSal)
kable(summaryTpHSal,caption="summary of data")



library(dplyr)
data <- data.frame(
  ID_SITE = c(5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5),
  DATE = as.Date(c("10/6/2020", "18/6/2020", "26/6/2020", "5/11/2020", "12/11/2020", "20/11/2020", "27/11/2020", "3/4/2021", "8/4/2021", "16/4/2021", "7/6/2021", "15/6/2021", "21/6/2021", "29/6/2021"), format = "%d/%m/%Y"),
  T = c(15.416, 15.892, 16.824, 13.417, 13.241, 13.048, 12.848, 11.072, 11.23, 10.971, 15.061, 16.295, 15.688, 15.399),
  S = c(34.733, 34.632, 34.704, 34.552, 34.424, 34.56, 34.31, 33.62, 33.8, 34.156, 34.626, 34.762, 34.865, 35.014),
  PH = c(8.138, 7.959, 7.953, 7.799, 8.055, 7.791, 7.794, 7.9, 7.93, 7.91, 7.94, 7.91, 7.93, 7.88)
)
data$DATE <- as.POSIXlt(data$DATE)
data$Month <- data$DATE$mon + 1
data$Year <- data$DATE$year + 1900
averagestemp_ph_S <- aggregate(cbind(T, S, PH) ~ Month + Year, data, FUN = mean)
june_2020 <- subset(averagestemp_ph_S, Month == 6 & Year == 2020)
june_2021 <- subset(averagestemp_ph_S, Month == 6 & Year == 2021)
kable(averagestemp_ph_S,caption="average temp, pH and salinity")



sd_temp <- aggregate(`T` ~ Month + Year, data, FUN = sd)
sd_pH <- aggregate(PH ~ Month + Year, data, FUN = sd)
sd_salinity <- aggregate(S ~ Month + Year, data, FUN = sd)
temp_sd <- merge(sd_temp, sd_salinity, by = c("Month", "Year"))
temp_sd <- merge(temp_sd, sd_pH, by = c("Month", "Year"))
kable(temp_sd, caption = "SD for each month")





```
averages2 <- aggregate(`Volume (mL)`/ `Abundance (n)` ~ Date, prey_ab_vol, FUN = mean)
print(averages2)

```{r, echo=FALSE}

#average volume per individual 

averagesvolume <- data.frame(
  "Month" = c(6, 11, 4, 6),
  "Year" = c(2020, 2020, 2021, 2021),
  "Volume" = c(3.8781924, 12.6688840, 3.0448443, 0.8122513)
)

kable(averagesvolume,caption= "average volume per sponge (mL)")

merged_dataset <- merge(averagesvolume, averagestemp_ph_S, by = c('Year', 'Month'))


````
correlation_tests <- list()

correlation_tests$Volume_T <- cor.test(merged_dataset$Volume, merged_dataset$T)
correlation_tests$Volume_S <- cor.test(merged_dataset$Volume, merged_dataset$S)
correlation_tests$Volume_PH <- cor.test(merged_dataset$Volume, merged_dataset$PH)

print(correlation_tests)

```{r, echo=FALSE}
correlation_table <- data.frame(
  "Metric" = c("pH", "Salinity", "Temperature"),
  "95 CI low" = c(-0.9877566, -0.9670712, -0.9777259),
  "95 CI high" = c(0.8798354, 0.9540408, 0.9324436),
  "Correlation" = c(-0.5262361, -0.08480915, -0.2758412),
  "p.value" = c(0.4738, 0.9152, 0.7242),
  "DF" = c(2, 2, 2)
)

kable(correlation_table, caption = "Pearsons product-moment correlation, abiotic factors on volume")


