---
title: "Hypothesis1"
output:
  pdf_document: default
  html_document: default
date: "2023-06-16"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = "https://cloud.r-project.org")

library(car)
library(lme4)
library(multcomp)
library(ggplot2)
install.packages("knitr")
install.packages("kableExtra")


library(knitr)
library(kableExtra)






```
Hypothesis 1: The size-correct abundance of H.perlevis is influenced by month

```{r, echo=FALSE}
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

```


```{r, echo=FALSE}
prey_ab_vol <- read.csv('Dataa/prey_ab_vol.csv')
prey_ab_vol$n <- prey_ab_vol$N.H..perlevis
prey_ab_vol$row_id <- factor(1:nrow(prey_ab_vol))
prey_ab_vol$Date <- factor(prey_ab_vol$Date)
prey_ab_vol <- prey_ab_vol[, colSums(is.na(prey_ab_vol)) == 0]

#cleaning up names 
# Rename specific columns by name
old_names <- c("N.quadrat", "N.H..perlevis","Volumen.Hym..mL."," Date","row_id")  # Example: column names to be renamed
new_names <- c("Quadrant number ", "Abundance (n)","Volume (mL)","Date","row_id")  # Example: new names for the columns

for (i in 1:length(old_names)) {
  colnames(prey_ab_vol)[colnames(prey_ab_vol) == old_names[i]] <- new_names[i]
}


#summary statistics 

# Summary staitistcs for dataset 1
summaryprey_ab_vol <- summary(prey_ab_vol)
summary_table1 <- as.data.frame(summaryprey_ab_vol)
filtered_table1 <- summary_table1[complete.cases(summary_table1), ]
kable(filtered_table1, align = "c", caption = "Summary for dataset 1")


# SD Table
sd_abundance <- sd(prey_ab_vol$`Abundance (n)`)
sd_volume <- sd(prey_ab_vol$`Volume (mL)`)

sd_table2 <- data.frame(
  Metric = c("Abundance", "Volume"),
  StandardDeviation = c(sd_abundance, sd_volume)
)
kable(sd_table2, align = "c", caption = "Standard deviation of Abundance (n) and Volume (mL)")

# mean for each date
meansdabundance <- aggregate(`Abundance (n)` ~ Date, data = prey_ab_vol, FUN = function(x) c(mean = mean(x), SD = sd(x)))
meansdvolume <- aggregate(`Volume (mL)` ~ Date, data = prey_ab_vol, FUN = function(x) c(mean = mean(x), SD = sd(x)))

summary_table <- merge(meansdabundance, meansdvolume, by = "Date", suffixes = c("Abundance", "Volume"))
kable(summary_table, align = "c", caption = " mean volume and abundance for dates")

# to find out when highest and lowest volumes were recorded for each month 

#volume
highest_volume <- max(prey_ab_vol$`Volume (mL)`)
highest_volume_rows <- prey_ab_vol[prey_ab_vol$`Volume (mL)` == highest_volume, ]
lowest_volume <- min(prey_ab_vol$`Volume (mL)`)
lowest_volume_rows <- prey_ab_vol[prey_ab_vol$`Volume (mL)` == lowest_volume, ]

merged_rows <- rbind(highest_volume_rows, lowest_volume_rows)
kable(merged_rows, align = "c", caption = "Highest and lowest volume (mL)")

#abundance

highest_abundance<- max(prey_ab_vol$`Abundance (n)`)
highest_abundance_rows <- prey_ab_vol[prey_ab_vol$`Abundance (n)` == highest_abundance, ]
lowest_abundance <- min(prey_ab_vol$`Abundance (n)`)
lowest_abundance_rows <- prey_ab_vol[prey_ab_vol$`Abundance (n)` == lowest_abundance, ]

merged_rows3 <- rbind(highest_abundance_rows, lowest_abundance_rows)
kable(merged_rows3, align = "c", caption = "Highest and lowest abundance (n)")


# GLM models

#first model with interaction term, overdispersed (n ~ ln.biovolume * Date + (1 | row_id)
prey_ab_vol$ln.biovolume <- log(prey_ab_vol$`Volume (mL)`/prey_ab_vol$`Abundance (n)`)
m1 <- glm(`Abundance (n)` ~ ln.biovolume*Date,family=poisson,data=prey_ab_vol)

#overdispersion function
table6 <-(overdisp_fun(m1))
kable(table6, align = "c", caption = "Overdispersion for GLM with interaction term")

#new model
# new model no overdispersion
m2 <- glmer(n ~ ln.biovolume*Date +(1|row_id),family=poisson,data=prey_ab_vol)
my <- m2
#m2: n ~ ln.biovolume * Date + (1 | row_id)
summary_tablea <- data.frame(
  `Fixed effects` = c(
    "(Intercept)", "ln.biovolume", "DateJun20", "DateJun21", "DateNov20", "ln.biovolume:DateJun20",
    "ln.biovolume:DateJun21", "ln.biovolume:DateNov20"
  ),
  Estimate = c(3.1512, -0.5586, 0.7878, 0.9128, -0.4735, 0.1806, -0.2845, 0.2049),
  `Std. Error` = c(0.2868, 0.2564, 0.6705, 0.5771, 0.3547, 0.3570, 0.5233, 0.3746),
  `z value` = c(10.988, -2.178, 1.175, 1.582, -1.335, 0.506, -0.544, 0.547),
  `Pr(>|z|)` = c("<2e-16 ***", "0.0294 *", "0.2400", "0.1137", "0.1819", "0.61302", "0.5867", "0.5843")
)
kable(summary_tablea, align = "c", caption = "Summary of Model 2, with interaction and row id")

# overdispersion
table7 <-(overdisp_fun(m2))
kable(table7, align = "c", caption = "Overdispersion for GLM model 2")


#chisq
table6.1 <-(drop1(m2,test='Chisq'))
kable(table6.1, align = "c", caption = "Chisq test with interaction model")

# no need for interaction term, new modellll

m3 <- glmer(`Abundance (n)` ~ ln.biovolume + Date +(1|row_id),family=poisson,data=prey_ab_vol)


fixed_effects <- data.frame(
  "Variable" = c("(Intercept)", "ln.biovolume", "DateJun20", "DateJun21", "DateNov20"),
  "Estimate" = c(3.0714, -0.4722, 1.0911, 0.5996, -0.4443),
  "Std. Error" = c(0.2128, 0.1428, 0.2997, 0.2171, 0.3093),
  "z value" = c(14.435, -3.307, 3.640, 2.762, -1.436),
  "Pr(>|z|)" = c("< 2e-16 ***", "0.000943 ***", "0.000272 ***", "0.005744 **", "0.150949")
)

kable(fixed_effects, align = "c", caption = "Summary of model 3")

#chisq
table6.2 <-(drop1(m3,test='Chisq'))
kable(table6.2, align = "c", caption = "M3: chisq")



library(knitr)
linear_hypotheses <- data.frame(
  Hypothesis = c("Jun20 - Apr21 == 0", "Jun21 - Apr21 == 0", "Nov20 - Apr21 == 0", 
                 "Jun21 - Jun20 == 0", "Nov20 - Jun20 == 0", "Nov20 - Jun21 == 0"),
  Estimate = c(1.0911, 0.5996, -0.4443, -0.4915, -1.5354, -1.0438),
  `Std. Error` = c(0.2997, 0.2171, 0.3093, 0.2685, 0.4569, 0.3071),
  `z value` = c(3.640, 2.762, -1.436, -1.831, -3.360, -3.399),
  `Pr(>|z|)` = c("0.00146 **", "0.02585 *", "0.44277", "0.23402", "0.00400 **", "0.00349 **")
)

kable(linear_hypotheses,caption = "Model 3:Linear Hypotheses Summary,Tukey")

library(knitr)

data <- data.frame(
  Variable = c("Jun20 - Apr21", "Jun21 - Apr21", "Nov20 - Apr21", "Jun21 - Jun20", "Nov20 - Jun20", "Nov20 - Jun21"),
  Estimate = c(1.09108, 0.59956, -0.44427, -0.49153, -1.53535, -1.04382),
  lwr = c(0.33509, 0.05204, -1.22451, -1.16881, -2.68788, -1.81853),
  upr = c(1.847083, 1.14707, 0.33598, 0.18576, -0.38282, -0.26912)
)

kable(data,caption= "confidence interval tukey test")

me <- (overdisp_fun(m3))
kable(me,caption= "m3 overdispersionfun")












ggplot(prey_ab_vol, aes(x = Date, y = sqrt(`Abundance (n)`))) +
  geom_boxplot(fill = "#FFDAB9", color = "black") +
  labs(title = "Relationship between abundance (n) per quadrant and month") +
  xlab("Date") +
  ylab("sqrt(abundance)")


pred <- predict(m3,type='response',re.form=NA)
ggplot(prey_ab_vol,aes(x=ln.biovolume,y=log(`Abundance (n)`),color=Date)) + geom_point() +
  geom_line(aes(y=log(pred))) +
    xlab("Log(Biomass)") +
  ylab("Log(Abundance)")




#levenes test

levene_results <- data.frame(
  "-" = c("group", ""),
  Df = c("3", "67"),
  `F value` = c(0.8323, NA),
  `Pr(>F)` = c(0.4808, NA)
)

kable(levene_results,caption= "levenes test on pearson residuals")


#residuals
residuals <- resid(m3, type = "pearson")
qqnorm(residuals)
qqline(residuals)
#random effects
qqnorm(ranef(m3)$row_id[,1]); qqline(ranef(m3)$row_id[,1])






