---
title: "NSCI7915 Preliminary Analysis" 
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2023-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, include= FALSE}

library(lme4)
library(multcomp)
````
<div align="center"><font size= 5>**Hypothesis 1: The abundance of H.perlevis is influenced by month**</div></font>

Metadata: Nudibranch predation boosts sponge silicon cycling (2023)

Independent Variable (IV) = Month (Jun20, Nov20, Apr20, Jun21)

Dependent variable (D) = Abundance (n)

Null Hypothesis (H0): There is no significant relationship between the abundance of Hymeniacidon perlevis and the month. 

Alternative Hypothesis (HA): There is a relationship between the abundance of Hymeniacidon perlevis and the month.

Due to the predator Doris Verrucosa and its peak abundance season being late June/early July we expect to see a relationship of June 2021 = June 2020 > April 2020 > November 2021 in regards to biomass. 



```{r, echo=FALSE}

prey_ab_vol <- read.csv('Dataa/prey_ab_vol.csv')
prey_ab_vol$row_id <- factor(1:nrow(prey_ab_vol))
Summaryprey_ab_vol <- summary(prey_ab_vol)
```

<div align = "center"><font size= 1.5>*Table 1: Statistical summary of abundance and volume data from the 2023 study*</div></font>

```{r, echo=FALSE}
print(Summaryprey_ab_vol)
```
<div align = "center"><font size= 1.5>*Table 2: Standard deviation of abundance and volume data from the 2023 study for hypothesis 1 & 2*</div></font>
```{r, echo=FALSE}
sd_N.H..perlevis <- sd(prey_ab_vol$N.H..perlevis)
sd_Volumen.Hym..mL. <- sd(prey_ab_vol$Volumen.Hym..mL.)

SDresults <- paste("Standard Deviation (N.H..perlevis):", sd_N.H..perlevis, "Standard Deviation (Volumen.Hym..mL.):", sd_Volumen.Hym..mL.)
print(SDresults)
  
boxplot(sqrt(N.H..perlevis)~ Date,prey_ab_vol,col="lemonchiffon", main = "Relationship between abundance (n) per quadrant and month")
````

<div align = "center"><font size= 1.5>*Figure 1: Boxplot of the relationship between abundance (n) per quadrant and the month the measurement was taken.(x-axis): The month the abundance was measured for and (y-axis): the square root of the abundance.*</div></font>


```{r, echo=FALSE}
m1 <- glm(N.H..perlevis ~ Date,prey_ab_vol,family='poisson')
m1 <- glmer(N.H..perlevis ~ Date + (1|row_id),prey_ab_vol,family='poisson')
````





<div align = "center"><font size= 1.5>*Table 3: Chisq drop1 test results*</div></font>
```{r, echo=FALSE}
drop1(m1,test='Chisq')
```
<div align = "center"><font size= 1.5>*Table 4: Generalised linear mixed effects model (GLMM) and posthoc tukey test*</div></font>
```{r, echo=FALSE}
summary(glht(m1, linfct = mcp(Date = "Tukey")))
````

```{r, echo=FALSE}
residualsm1 <- residuals(m1)
qqnormm1 <- qqnorm(residualsm1)
qqnormm1 <- qqline(residualsm1)

```
<div align = "center"><font size= 1.5>*Figure 2: QQ plot for GLMM model*</div></font>

# poison distribution histogram? qq plot? idk

<div align = "center"><font size= 5>**Results**</div></font>

A boxplot was used the visualise the relationship between the abundance of H. perlevis over the months. The single terms deletions analysis was used to assess the significance of the ‘date’ variable in predicting the volume of H.perlevis.  The initial model fitted with the variable ‘date’ as the predictor had an Akaike information criterion (AIC) of 574.87 and npara of 3. Removing ‘date’ from the model slightly increased AIC to 574.94. The likelihood ratio test (LRT) statistic was 6.0749, with a 0.108 p-value. The removal of the ‘date’ variable did not significantly impact the model fit. The ‘date’ variable may not be a significant predictor of the abundance (n) of H.perlevis. Results from the Generalized Linear Mixed-effects Model (glmer) and posthoc Tukey test showed there was not sufficient evidence (p-value <0.05) to reject the null hypothesis while accounting for the random effects;Jun20-Apr 21 (Pr= 0.2826), Jun21-Apr21 (Pr = 0.0901), Nov20-Apr21 (Pr = 0.8895), Jun21-Jun20 (Pr =0.9179), Nov20-Jun20 (Pr=0.7895), Nov20-Jun21 (Pr =0.4547). 
 


<div align = "center"><font size= 5>**Assumptions of Analyses **</div></font>



**Independence of observations**: No two observations in the dataset are related to each other or affect each other in any way.

**Poisson distribution**: This distribution and its associated link function (log) has been selected for the data when using the GLMM model.

**Linearity**: The relationship between Date and the log of the abundance (n) is linear on the log scale

**Homogeneity of variance**: Poisson distribution assumed that the variance of the abundance is equal across all levels for the variable date.

**Random effects assumption**: The random intercept term ‘1|row_id’ was used in the model allowing for the intercepts in each row to vary randomly.






<div align="center"><font size= 5>**Hypothesis 2: Does the month have a effect on the volume of H.perlevis?**</div></font>

Metadata from: Nudibranch predation boosts sponge silicon cycling (2023)

IV= Month (Jun20,Nov20,Apr20,Jun21)

D= H.perlevis volume (mL)

Null Hypothesis (H0): There is no significant relationship between the volume of Hymeniacidon perlevis and the month

Alternative Hypothesis (HA): There is a significant relationship between the volume of Hymeniacidon perlevis and the month

Analyses: Pearson correlation test, one-way ANOVA, Tukey post-hoc test, Boxplot, levenes test



<div align = "center"><font size= 1.5>*Table 5: Pearsons correlation between the log of abundance (n) and volume (mL) *</div></font>

```{r, echo=FALSE}

cor.test(log(prey_ab_vol$N.H..perlevis),log(prey_ab_vol$Volumen.Hym..mL.))
````

<div align = "center"><font size= 1.5>*Table 6: One way anova and posthoc tukey test between the volume (mL) of H.perlevis and the months *</div></font>

```{r, echo=FALSE}
m7 <- aov(log(Volumen.Hym..mL.) ~ Date,prey_ab_vol)
anova(m7)
TukeyHSD(m7,ordered=TRUE)
````

```{r, echo=FALSE}
boxplot(log(Volumen.Hym..mL.)~Date,prey_ab_vol,main = "Relationship between volume and month",col= 'pink')
````
<div align = "center"><font size= 1.5>*Figure 3: Boxplot of the relationship between month/date and the log of the volume (mL) for H.perlevis.(x axis): The month the volume was measured and (y axix): the log of the volume.*</div></font>





<div align = "center"><font size= 1.5>*Table 7: Levenes test for homogeneity of variance  *</div></font>
```{r, echo=FALSE}
library(car)
leveneTest(m7)
```

  
```{r, echo=FALSE}
residualsm5 <- residuals(m7)
qqnormm5 <- qqnorm(residualsm5)
qqnormm5 <- qqline(residualsm5)
```
<div align = "center"><font size= 1.5>*Figure 4: QQ plot for ANOVA model *</div></font>





<div align = "center"><font size= 5>**Results **</div></font>


A Pearson correlation test was done between the abundance of H.perlevis (n) and the volume of H.perlevis (mL). The correlation coefficient was 0.5149891 indicating a moderate positive linear relationship (df= 69, p-value = 4.326e-06, confidence interval = 95%). This analysis provides evidence to reject the null hypothesis. The ANOVA analysis indicated a significant effect of the variable ‘date’ on the logarithmically transformed variable ‘volume of H.perlevis’ (F value (3,67) = 51.380, p value= 2.35e-17 ((<0.05)). A post hoc Tukey test showed that the volume of H.perlevis among the months differed significantly at p <.05 and confidence level of 95% [April 21- Nov 20 (p value= 3.2e-4), Jun 21 – Nov 20 (p value= 0), Jun 20- Nov 20 (p value= 0), Jun 21 – Apr 21 (p-value = 4.4e-2), Jun 20- Apr 21 (p value= 0), Jun 20 – Jun 21 (p value= 5.7e-6). The results of the ANOVA test support the alternative and reject the null hypothesis. Boxplot no. Levene’s test was conducted to examine the homogeneity of variables across the levels for the independent variable ‘date’. A non-significant result was observed (F value (3,67) = 0.4505, p value= 0.7178). There is no evidence to reject the assumption of homogeneity of variance.

<div align = "center"><font size= 5>**Assumptions of Analyses **</div></font>


**Independence of observations**: No two observations in the dataset are related to each other or affect each other in any way.


**Normal distribution**: The data is normally distributed (refer to figure 4)


**Linearity**: The relationship between Date and the volume of H.perlevis is linear 


**Homogeneity of variance**: The variance of the volume of H.perlevis is equal across all groups for the months (tested with levenes test)









<div align="center"><font size= 5>**Hypothesis 3: Is there a relationship between average volume of H.perlevis per month and pH, salinity and temperature averages for the months**</div></font>

Metadata from: Nudibranch predation boosts sponge silicon cycling (2023)
Temp,Salinity and pH values obtained from Somlit.

IV= average Temp (T), Salinity (S) and PH values for the month the volume was measured

D=  average sponge volume per month (mL sponge m-2) given by ((ponge volume mL./abundance) per quadrant, averages for the months)

Null Hypothesis (H0): There is no signfiicant relationship between the average volume of H.perlevis per month (D) and the average temperature (T), salinity (S), and pH values for the corresponding months.

Alternative Hypothesis (HA): There is a signfiicant relationship between the average volume of H.perlevis per month (D) and the average temperature (T), salinity (S), and pH values for the corresponding months.

Analyses: pearsons correlation

<div align = "center"><font size= 1.5>*Table 8: Summary statistics for temperature, salinity and pH data from SOMLIT *</div></font>

```{r, echo=FALSE}

suppressPackageStartupMessages({
  library(dplyr)
})

TpHSal <- read.csv('Dataa/TempsalinitypH.csv')
empty_columns <- which(colSums(is.na(TpHSal)) == nrow(TpHSal))
TpHSal <- subset(TpHSal, select = -empty_columns)
summaryTpHSal <- summary(TpHSal)
print(summaryTpHSal)

sdT <- sd(TpHSal$T)
sdpH <- sd(TpHSal$PH)
Sal <- sd(TpHSal$S)
SDresults2 <- paste("Standard Deviation (T):", sdT, "Standard Deviation (pH):", sdpH, "Standard Deviation (Sal):", Sal)
print(SDresults2)

```
<div align = "center"><font size= 1.5>*Table 9: Averages data for Temp, pH, salinity and volume per month brackets *</div></font>


```{r, echo=FALSE}
library(dplyr)
data <- data.frame(
  ID_SITE = c(5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5),
  DATE = as.Date(c("10/6/2020", "18/6/2020", "26/6/2020", "5/11/2020", "12/11/2020", "20/11/2020", "27/11/2020", "3/4/2021", "8/4/2021", "16/4/2021", "7/6/2021", "15/6/2021", "21/6/2021", "29/6/2021"), format = "%d/%m/%Y"),
  T = c(15.416, 15.892, 16.824, 13.417, 13.241, 13.048, 12.848, 11.072, 11.23, 10.971, 15.061, 16.295, 15.688, 15.399),
  S = c(34.733, 34.632, 34.704, 34.552, 34.424, 34.56, 34.31, 33.62, 33.8, 34.156, 34.626, 34.762, 34.865, 35.014),
  PH = c(8.138, 7.959, 7.953, 7.799, 8.055, 7.791, 7.794, 7.9, 7.93, 7.91, 7.94, 7.91, 7.93, 7.88)
)
data$DATE <- as.POSIXlt(data$DATE)
data$Month <- data$DATE$mon + 1
data$Year <- data$DATE$year + 1900
averagestemp_ph_S <- aggregate(cbind(T, S, PH) ~ Month + Year, data, FUN = mean)
june_2020 <- subset(averagestemp_ph_S, Month == 6 & Year == 2020)
june_2021 <- subset(averagestemp_ph_S, Month == 6 & Year == 2021)
print(averagestemp_ph_S)
averages2 <- aggregate(Volumen.Hym..mL. / N.H..perlevis ~ Date, prey_ab_vol, FUN = mean)
print(averages2)
averagesvolume <- data.frame(
  Month = c(6, 11, 4, 6),
  Year = c(2020, 2020, 2021, 2021),
  Volumen.Hym..mL..N.H..perlevis = c(3.8781924, 12.6688840, 3.0448443, 0.8122513)
)
```

<div align = "center"><font size= 1.5>*Table 10: Correlation analysis for volume of H.perlevis vs T,S and pH *</div></font>


```{r, echo=FALSE}
merged_dataset <- merge(averagesvolume, averagestemp_ph_S, by = c('Year', 'Month'))
correlation2 <- cor(merged_dataset[, c("Volumen.Hym..mL..N.H..perlevis", "T", "S", "PH")])
print(correlation2)

```




<div align = "center"><font size= 5>**Results **</div></font>

A correlation test was used to show the relationship between volume of H.perlevis for the different months and T, S and Ph for the corresponding month. Volume had a weak negative correlation with temperature (r= -0.276) and a weak negative correlation with salinity (r= -0.085). It has a moderate negative correlation with the variable pH (r= -0.526). 


<div align = "center"><font size= 5>**Assumption of Analyses **</div></font>


**Independence of observations**: No two observations in the dataset are related to each other or affect each other in any way.

**Linearity**: The relationship between Date and the volume of H.perlevis is linear 

**Homogeneity of variance**: The variance of the volume of H.perlevis is equal across all groups for the months. 




<div align="center"><font size= 5>**Hypothesis 4: What is the relationship between the abundance of H.perlevis and biomass per haibitat (2022 study)**</div></font>

metadata: Sponge contribution to the silicon cycle of a diatom-rich shallow bay (2022) 
IV= habitat& abundance (N sponge m-2)
D= biomass (mL sponge m-2)
Habitats= Rocky intertidal,Rocky subtidal,Maerl,Shallow muds,Heterog seds,Circ.coarse seds


Null Hypothesis (H0): There is no significant relationship between biomass and the interaction between Habitat and abundance

Alternative Hypothesis (HA): There is a significant relationship between biomass and the interaction between Habitat and abundance

Analyses: scatterplot, correlation test, linear regression analysis, anova (m10b and m10), regression analysis

```{r, echo=FALSE}
library(ggplot2)
data2022 <- read.csv('Dataa/data2022.csv')
data2022$row_id <- factor(1:nrow(data2022))
data2022 <- data2022[data2022$Abundance.sponges > 0, ]
data2022$Sponge.biomass <- as.numeric(gsub("[^0-9.]", "", data2022$Sponge.biomass))
```

<div align = "center"><font size= 1.5>*Table 11: Summary statistics for abundance (N sponge m-2) and biomass (mL sponge m-2)  of H.perlevis sponges from 2022 study *</div></font>

```{r, echo=FALSE}
summarydata2022 <- summary(data2022)
print(summarydata2022)
sdabundance <- sd(data2022$Abundance.sponges)
sdbiomass <- sd(data2022$Sponge.biomass)

sd2022 <- paste("Standard Deviation (Abundance):", sdabundance, "Standard Deviation (Biomass):", sdbiomass)
print(sd2022)
```

```{r, echo=FALSE}
data2022$log.n <- log(data2022$Abundance.sponges)
data2022$log.biomass <- log(data2022$Sponge.biomass)

##normal plot
plot(data2022$log.n, data2022$log.biomass, xlab = "log abundance (N sponge m-2) ", ylab = "log Biomass (mL sponge m-2)", main = "Relationship between Abundance and Biomass",type = "n")

j <- 0
for (i in unique(data2022$Habitat)) {
  points(data2022$log.n[data2022$Habitat == i], data2022$log.biomass[data2022$Habitat == i],
         pch = j + 1, col = c("lightpink1", "lightblue", "mediumpurple1","orchid1", "palevioletred1",'lightgreen')[j + 1])
  j <- j + 1
}

legend("bottomright", legend = unique(data2022$Habitat),
       pch = 1:length(unique(data2022$Habitat)),
       col = c("lightpink1", "lightblue", "mediumpurple1", "orchid1", "palevioletred1", 'lightgreen'),
       title = "Habitat", horiz = TRUE, cex = 0.35, pt.cex = 1.2, pt.bg = "white")
```
<div align = "center"><font size= 1.5>*Figure 5: Scatterplot of the relationship between (x): log abundance (N sponge m-2) and (y): log biomass (mL sponges m-2), per habitat.*</div></font>

```{r, echo=FALSE}
##ggplot

ggplot(data2022, aes(x = log.n, y = log.biomass, color = Habitat)) +
  geom_point() +
  labs(x = "log abundance (N sponge m-2)", y = "log Biomass (mL sponge m-2)") +
  theme_minimal() +
  facet_wrap(~ Habitat, nrow = 2, scales = "free") +
  scale_color_manual(values = c("lightpink1", "lightgreen", "paleturquoise", "orchid1", "mediumpurple1", "palevioletred1"),
                     breaks = unique(data2022$Habitat),
                     labels = unique(data2022$Habitat),
                     name = "Habitat")

````

<div align = "center"><font size= 1.5>*Figure 6: Scatterplots of the relationship between (x): log abundance (N sponge m-2) and (y): log biomass (mL sponges m-2), per habitat. Representation 2 using ggplot plugin.*</div></font>



<div align = "center"><font size= 1.5>*Table 12: Correlation analysis between abundance and biomass of H.perlevis. For hypothesis 4. *</div></font>
```{r, echo=FALSE}
correlation3 <- cor(data2022$Abundance.sponges, data2022$Sponge.biomass)
correlation3
````
<div align = "center"><font size= 1.5>*Table 13: Anova analysis of two different linear models between the log abundance and log biomass of H.perlevis per habitat. M1 considers a interaction between habitat and abundance, while M2 does not. For hypothesis 4. *</div></font>
```{r, echo=FALSE}
m10 <- lm(log.biomass ~ Habitat*log.n, data = data2022)
m10b <- lm(log.biomass ~ Habitat + log.n, data = data2022)
anova(m10b,m10)
````
<div align = "center"><font size= 1.5>*Table 14: Summary of M1*</div></font>
```{r, echo=FALSE}
summary(m10b)
````




```{r, echo=FALSE}

residualsm3 <- residuals(m10)
qqnormm3 <- qqnorm(residualsm3)
qqnormm3 <- qqline(residualsm3)

````
<div align = "center"><font size= 1.5>*Figure 7: QQ plot for ANOVA model in hypothesis 4*</div></font>






<div align = "center"><font size= 5>**Results **</div></font>


A correlation test was used to show the relationship between the variables
The correlation coefficient between the variables abundance and biomass. A correlation coefficient of 0.1208 was received, indicating a moderate positive linear relationship. A scatterplot was created to illustrate the relationship between abundance and biomass per habitat. An ANOVA was used to compare two GLM models examining the relationship between biomass and the interaction between habitat and abundance (F statistic= 1.1137 P-value= 0.3562). Model 1 (M1) included the predictors habitat and abundance, while Model 2 (M2) included an interaction term between habitat and abundance. M1 (Res.Df= 135, RSS= 196.2390), M2 ( Res.DF= 130, RSS= 188.1783). There is no significant difference between M1 and M2.  The linear regression analysis of M1 (Residual standard error: 1.206 on 135 degrees of freedom) showed that the variables abundance and habitat have significant effects on the biomass of H.perlevis (F value(135)= 26.75, p-value= 2.2e-16)). The model explains approx. 54.31% of the variance in the biomass variable (multiple R-squared). There is strong evidence to support the alternative hypothesis and reject the null.

<div align = "center"><font size= 5>**Assumption of Analyses **</div></font>


**Independence of observations**:No two observations in the dataset are related to each other or affect each other in any way.

**Normal distribution**: The data is normally distributed (represented by figure 7).

**Linearity**: The relationship between abundance and biomass of H.perlevis is linear 

**Homogeneity of variance**: The variance of the volume of H.perlevis is equal across all groups for the habitats.


