---
title: "help"
output: word_document
date: "2023-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hypothesis 1: The relationship between abundance of H.perlevis and the months (from 2023 study) 
independent: Month of study, Dependent: abundance 
Analysis: Glmer model including a random effect, modelling the variability among different levels, chi-squared test, post-hoc tukey analysis

```{r}
prey_ab_vol <- read.csv('Dataa/prey_ab_vol.csv')
prey_ab_vol$row_id <- factor(1:nrow(prey_ab_vol))

boxplot(sqrt(N.H..perlevis)~ Date,prey_ab_vol,col="lemonchiffon", main = "Relationship between abundance and month")

m1 <- glm(N.H..perlevis ~ Date,prey_ab_vol,family='poisson')
m1 <- glmer(N.H..perlevis ~ Date + (1|row_id),prey_ab_vol,family='poisson')
drop1(m1,test='Chisq')
summary(glht(m1, linfct = mcp(time = "Tukey")))
```

Hypothesis 2: The relationship between  volume of H.perlevis and the months (2023 study)

```{r}

cor.test(log(prey_ab_vol$N.H..perlevis),log(prey_ab_vol$Volumen.Hym..mL.))

m1 <- aov(log(Volumen.Hym..mL.) ~ Date,prey_ab_vol)
anova(m1)
TukeyHSD(m1,ordered=TRUE)
boxplot(log(Volumen.Hym..mL.)~Date,prey_ab_vol,main = "Relationship between volume and month",col= 'pink')

leveneTest(m1)
```

Hypothesis 3: The relationship between average volume of H.perlevis per month and pH, salinity and temperature averages for the months (2022)

```{r}
TpHSal <- read.csv('Dataa/TempsalinitypH.csv')
TpHSal$row_id <- factor(1:nrow(TpHSal))
library(dplyr)
data <- data.frame(
  ID_SITE = c(5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5, 5),
  DATE = as.Date(c("10/6/2020", "18/6/2020", "26/6/2020", "5/11/2020", "12/11/2020", "20/11/2020", "27/11/2020", "3/4/2021", "8/4/2021", "16/4/2021", "7/6/2021", "15/6/2021", "21/6/2021", "29/6/2021"), format = "%d/%m/%Y"),
  T = c(15.416, 15.892, 16.824, 13.417, 13.241, 13.048, 12.848, 11.072, 11.23, 10.971, 15.061, 16.295, 15.688, 15.399),
  S = c(34.733, 34.632, 34.704, 34.552, 34.424, 34.56, 34.31, 33.62, 33.8, 34.156, 34.626, 34.762, 34.865, 35.014),
  PH = c(8.138, 7.959, 7.953, 7.799, 8.055, 7.791, 7.794, 7.9, 7.93, 7.91, 7.94, 7.91, 7.93, 7.88)
)
data$DATE <- as.POSIXlt(data$DATE)
data$Month <- data$DATE$mon + 1
data$Year <- data$DATE$year + 1900
averagestemp_ph_S <- aggregate(cbind(T, S, PH) ~ Month + Year, data, FUN = mean)
june_2020 <- subset(averagestemp_ph_S, Month == 6 & Year == 2020)
june_2021 <- subset(averagestemp_ph_S, Month == 6 & Year == 2021)
print(averagestemp_ph_S)
averages2 <- aggregate(Volumen.Hym..mL. / N.H..perlevis ~ Date, prey_ab_vol, FUN = mean)
print(averages2)
averagesvolume <- data.frame(
  Month = c(6, 11, 4, 6),
  Year = c(2020, 2020, 2021, 2021),
  Volumen.Hym..mL..N.H..perlevis = c(3.8781924, 12.6688840, 3.0448443, 0.8122513)
)
merged_dataset <- merge(averagesvolume, averagestemp_ph_S, by = c('Year', 'Month'))
correlation2 <- cor(merged_dataset[, c("Volumen.Hym..mL..N.H..perlevis", "T", "S", "PH")])
```

Hypothesis 4: The relationship between abundance of H.perlevis and biomass per haibitat (2022 study)

````{r}
data2022 <- read.csv('Dataa/data2022.csv')
data2022$row_id <- factor(1:nrow(data2022))
data2022 <- data2022[data2022$Abundance>0,]
data2022$log.n <- log(data2022$N)
data2022$log.biomass <- log(data2022$biomass)

plot(data2022$log.n, data2022$log.biomass, xlab = "Abundance", ylab = "Biomass", main = "Relationship between Abundance and Biomass", type = 'n')

j <- 0
for (i in unique(data2022$Habitat)) {
  points(data2022$log.n[data2022$Habitat == i], data2022$log.biomass[data2022$Habitat == i],
         pch = j + 1, col = c("pink", "blue", "purple","green", "cyan",'grey')[j + 1])
  j <- j + 1
}

legend("bottomright", legend = unique(data2022$Habitat),
       pch = 1:length(unique(data2022$Habitat)),
       col = c("pink", "blue", "purple", "green", "cyan", 'grey'),
       title = "Habitat", horiz = TRUE, cex = 0.3, pt.cex = 1, pt.bg = "white")

correlation3 <- cor(data2022$log.n, data2022$log.biomass)
correlation3

m10 <- lm(log.biomass ~ Habitat*log.n, data = data2022)
m10b <- lm(log.biomass ~ Habitat + log.n, data = data2022)
anova(m10b,m10)
summary(m10)



````



